---
title: "Modeling Environmental Stochasticity"
author: "Eric R. Scott"
date: "2022-04-21"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
    highlight: kate
    theme:
      version: 4
      bootswatch: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(targets)
library(mgcv)
library(gratia)
library(bbmle)
library(tictoc)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

*Last compiled: `r Sys.Date()`*

# Purpose

Try out different ways of doing random effects from these sources:

-   <https://fromthebottomoftheheap.net/2021/02/02/random-effects-in-gams/>

-   Pedersen, Eric J., David L. Miller, Gavin L. Simpson, and Noam Ross.
    "Hierarchical Generalized Additive Models in Ecology: An Introduction with Mgcv." PeerJ 7 (May 27, 2019).
    <https://doi.org/10.7717/peerj.6876>.

# Load Data

```{r data, echo=TRUE}
tar_load(data_full)
```

```{r}
df <- data_full %>%
  filter(sdlg_prev == FALSE, habitat == "1-ha") %>%
  filter(!is.na(log_size_prev)) %>% 
  select(-L, -spei_history)
df
```

# Fit GAMs

With no random effect

```{r}
tic()
m0 <- bam(
  surv ~ s(log_size_prev, bs = "cr", k = 20),
  family = binomial,
    data = df,
    method = "fREML"
)
toc()
draw(m0)
df <- df %>% mutate(fitted0 = predict(m0, type = "response"))
```

With random intercepts for each year

```{r}
tic()
m1 <- bam(
  surv ~ s(log_size_prev, bs = "cr", k = 20) +
    s(year_fct, bs = "re"),
  family = binomial,
    data = df,
    method = "fREML"
)
toc()
draw(m1)
df <- df %>% mutate(fitted1 = predict(m1, type = "response"))
```

With random intercept and "slope" for each year.
Not clear to me what slope means exactly in this context.

```{r}
tic()
m2 <- bam(
  surv ~ s(log_size_prev, bs = "cr", k = 20) +
    s(year_fct, log_size_prev, bs = "re"),
  family = binomial,
    data = df,
    method = "fREML"
)
toc()
draw(m2)
df <- df %>% mutate(fitted2 = predict(m2, type = "response"))
```

With a global smoother and a different shapes for each year (but not different wiggliness).

```{r}
tic()
m3 <- bam(
  surv ~ s(log_size_prev, bs = "cr", k = 20, m = 2) +
    s(log_size_prev, year_fct, bs = "fs",  m = 2),
  family = binomial,
    data = df,
    method = "fREML"
)
toc()
draw(m3)
df <- df %>% mutate(fitted3 = predict(m3, type = "response"))
```

```{r}
AICtab(m0, m1, m2, m3)
```

The last one fits the data best, but also takes the longest to fit.
m2 is not far off with a dAIC of 5.

This is what the fitted values look like

```{r}
p0 <- ggplot(df, aes(x = log_size_prev)) +
  geom_line(aes(y = fitted0)) +
  labs(title = "no random effect")
p1 <- ggplot(df, aes(x = log_size_prev)) +
  geom_line(aes(y = fitted1, color = year_fct)) +
  labs(title = "random intercept")
p2 <- ggplot(df, aes(x = log_size_prev)) +
  geom_line(aes(y = fitted2, color = year_fct)) +
  labs(title = "random intercept and slope")
p3 <- ggplot(df, aes(x = log_size_prev)) +
  geom_line(aes(y = fitted3, color = year_fct)) +
  labs(title = "random shape")

library(patchwork)
p0 + p1 + p2 + p3 & labs(y = ".fitted") & theme(legend.position = "none") & plot_annotation(title=  "survival probability")
```
