---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    # template: AmNat_MS_template.tex
    # documentclass: scrarticle
    documentclass: article
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 12pt
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        \usepackage[sc]{mathpazo} %Like Palatino with extensive math support
        \usepackage[utf8]{inputenc}
        \usepackage{parskip}
        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
        \setlength{\parindent}{0pt}
    
        
#\sectionfont{\color{black}\fontsize{12}{16.8}\selectfont}
#\subsectionfont{\color{black}\fontsize{12}{16.8}\selectfont}
        
# number after font size is the interline spacing. It should about 20-25 % more 
# than the font size (for a 10 pt document, you can measure it is equal to 12 pt).
# \titleformat{\subsection}
# \titleformat*{\subsection}{\fontsize{12{16.8}\selectfont
        #     {\sffamily\Large\bfseries}{\thesection}{1em}{}[{\titlerule[0.8pt]}]

# When ready to submit, these will change to double space and add a 
# paragraph indent
# \usepackage{setspace}\doublespacing
# \setlength\parindent{22pt}

fig-cap-location: bottom
tbl-cap-location: top
geometry:
  - top=1in
  - left=1in
  - bottom=1in
  - right=1in
keywords: 
  - integral projection models
  - demography
  - population dynamics
  - environmental stochasticity
  - lagged effects
  
Notes: >
  Manuscript elements: Figure~1, figure~2, table~1, appendix~A.  
  Manuscript type: e-note


# abstract: >
#  The abstract should have 150 words. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(grateful)
library(tidyverse)
library(kableExtra)
library(here)
```

```{r}
#load targets
withr::with_dir(here(), {
  tar_load(c(
    aic_tbl_df,
    lambda_table,
    fig_pop_states
  ))
})
```

<!-- The actual document text starts here: -->

\pagebreak

# Abstract

<!-- AmNat Note Abstracts are no longer than 150 words. -->

Text of 150 words max summarizing this amazing paper.

\bigskip

**Keywords:** demography, environmental stochasticity, integral projection models, lagged effects, structured population models, population dynamics

\bigskip

**Manuscript elements:** Figure\~1, figure\~2, table\~1, appendix\~A..

\bigskip

**Manuscript type:** e-note

\pagebreak

<!-- AmNat Notes: communicate concise points, using either data or theory. Like Major Articles, they present insights of broad general significance and interest. Notes, on average, should be no more than 3000 words of text (not including the Literature Cited) and have no more than three figures and/or tables in print. -->

## Introduction

Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011]. Mathematical and statistical advances [e.g., @brooksStatisticalModelingPatterns2019; @williams2012avoiding]have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @reesEvolvingIntegralProjection2016; @ellnerDatadrivenModellingStructured2016]. Despite this progress, however, several important biological processes have proven challenging to incorporate in structured population models. In some cases this is because theoretical and analytical methods for doing so remain underdeveloped; in others this is because data with which to parametertize and assess alternative model structures are lacking [@ellnerDatadrivenModellingStructured2016; @metcalfStatisticalModellingAnnual2015].\

One of these biological processes is *Delayed Life-history Events* (i.e., DLHEs), also known as *Lagged Effects* [@beckermanPopulationDynamicConsequences2002]. Lagged effects are those in which the demographic vital rates observed in a given year are influenced -- or even determined by -- past environmental conditions. For instance, environmental conditions during juvenile development can shape the expression of traits (e.g., defensive spikes on *Daphnia*) that determine adult survival. Alternatively, the physiological mechanisms responsible for a vital rate can take an extended period of time to complete [@eversLaggedDormantSeason2021]; for example, flowering bud formation may be initiated several months before flowers appear [@crileyYearProductionHigh1994]. Vital rates can even be influenced by environmental conditions during the parental life-cycle or historical trade-offs between vital rates (e.g., delayed costs of reproduction, competition-colonization trade-offs).\

Although such lagged effects could potentially have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002], their their demographic impacts remain poorly understood [but see @molowny-horasChangesNaturalDynamics2017; @tenhumbergTimelaggedEffectsWeather2018; @williamsLifeHistoryEvolution2015]. There are two primary reasons for this limited understanding. The first is a lack of empirical data [sensu @eversLaggedDormantSeason2021]. Detecting lagged effects requires long-term data on both the putative lagged effects and their potential drivers [@metcalfStatisticalModellingAnnual2015], and studies to detect them can be challenging to design and maintain [@kussEvolutionaryDemographyLonglived2008a]. The second is challenge is technical - incorporating complex biological processes such as lagged effects in demographic models can render the models less tractable. <!-- would be good to close with a sentence stating that if model outputs are incorrect, thenm failure to do so could lead to major problems... -->\

We used a decade of survey and climate data to assess the effects of precipitation extremes on the demographic vital rates of the Amazonian understory herb *Heliconia acuminata* [@scottDelayedEffectsClimate2022]. Our analyses revealed that the effects of climate on vital rates could be delayed up to 36 months, and that the presence and duration of these effects could differ by vital rate and habitat type (i.e., continuous forest vs. forest fragments). Here we investigate how including lagged effects in Integral Projection Models influences projections of population growth rate (i.e., $\lambda$) and structure. To do so we parameterized and compared three different classes of Integral Projection Models: a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of SPEI on vital rates. Based on previous studies [@brunaArePlantPopulations2003; @brunaDemographicEffectsHabitat2005; @brunaExperimentalAssessmentHeliconia2002; @brunaHabitatFragmentationDemographic2002] and demographic theory [@caswell2001; @tuljapurkarPopulationDynamicsVariable1990] we predicted that:\

(i) $\lambda$ would be higher in forest than fragments regardless of model type,
(ii) that projections of $\lambda$ from deterministic models would be higher than those of stochastic models,
(iii) that $\lambda$ would be lowest for models including lagged effects, and
(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regardless of whether models included stochasticity or lagged effects.

## Methods

### *Study System and Demographic Data*

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot [@kressDiversityDistributionHeliconia1990] that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990]. While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H. acuminata* - grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010]. Understory *Heliconia* species typically produce fewer flowers and are pollinated by traplining hummingbirds [@stoufferForestFragmentationSeasonal1996; @brunaHeliconiaAcuminataReproductive2004].\

The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H. acuminata* found at Brazil's Biological Dynamics of Forest Fragments Project (BDFFP), located \~70 km north of Manaus, Brazil. The BDFFP reserves include both continuous forest and forest fragments that range in size from 1-100 ha. These fragment reserves were originally isolated in the early 1980’s by the creation of cattle pastures, with the secondary growth surrounding them periodically cleared to ensure their continued isolation. The habitat in all sites is non-flooded lowland rain forest with rugged topography. A complete summary of the BDFFP and its history can be found in @bierregaardLessonsAmazoniaEcology2001.\

In 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP's Continuous Forest Reserves (N=6 plots) and 1-ha Fragments (N=4 plots) in which all of the *Heliconia acuminata* were marked and measured. The plots were censused annually, at which time a team maked marked and measured new seedlings, identified any previously marked plants that died, and recorded the size of surviving individuals. Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants; in our site *H. acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = \[201, 1549\]; 1-ha median = 339, range = \[297, 400\]). A complete description of the demographic methods, data, and analyses to date can be found in [@brunaDemographyUnderstoryHerb2023].\

### *Construction of Integral Projection Models*

In preliminary investigation, we found that the survival and growth of plants was better explained by treating seedlings and mature plants separately. Seedlings are physiologically different from small plants because they necessarily lack the underground reserves (of carbohydrates and meristems) that a small, mature plant may have. Therefore, we used general IPMs to model population dynamics with seedlings treated as a separate discreet class not structured by size. General IPMs allow for combinations of continuous and discrete states and transitions between them [@ellnerDatadrivenModellingStructured2016].\

We built three classes of IPMs for comparison which each required different functional forms of their underlying vital rates models. The simplest IPM was a general, density-independent, deterministic IPM with four sub-kernels: growth and survival ($P$, @eq-P), fecundity ($F$, i.e. production of new seedlings, @eq-F), probability of staying a seedling (always 0), and recruitment ($R$, i.e. seedling survival and establishment, @eq-R) ([@eq-mature; @eq-sdlg], @fig-lifecycle). The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.\

$$
n(z^{\prime},t+1) = R(z^\prime)n_s(t) + \int_L^U P(z^\prime,z) n(z,t)\;dz
$$ {#eq-mature}

$$
n_s(t+1) = \int_L^U F(z) n(z,t) \; dz
$$ {#eq-sdlg}

$$
R(z^\prime) = s_sG_s(z^\prime)
$$ {#eq-R}

$$
P(z^\prime, z) = s(z)G(z^\prime, z)
$$ {#eq-P}

$$
F(z) = p_f(z)f(z)g
$$ {#eq-F}

The number and size of mature plants in the next census is determined by seedlings entering the mature plant population (i.e. recruitment <!--# *Is* this recruitment? -->) and survival and growth (or regression) of mature plants ([@eq-mature]). Seedlings survive ($s_s$) and grow into mature plants of a particular size ($G_s(z^\prime)$) ([@eq-R]).\

Mature plants survive as a function of size ($s(z)$), and grow (or regress) to a new size as a function of their previous size ($G(z^\prime, z)$) ([@eq-P]). Mature plants flower with a probability that is a function of size ($p_f(z)$) and produce a number of seeds as a function of size ($f(z)$), which germinate and establish as seedlings with probability $g$ ([@eq-F]).\

Vital rate models for growth ($G_s(z^\prime)$ and $G(z^\prime, z)$) , survival ($s_s$ and $s(z)$), and flowering ($p_f(z)$) were fit using the long term demographic dataset. For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package[@mgcv] in `r version$version.string` [@rcoreteam2020] <!--# probably include a supplement with detail on things like basis, number of knots, etc. -->. For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. intercept only models). For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple Gaussian model.\

To estimate reproduction we drew on additional data sources to estimate the number of fruits per flowering plant as a function of plant size and the number of seeds per fruit (together $f(z)$). Germination and establishment rates in continuous forest and forest fragments were estimated using data from .... <!--# Emilio: add more info here -->.\

To build the general, density-independent, stochastic, kernel-resampled IPMs, we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (@fig-lifecycle). The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years. The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly. This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.\

For the third method, we modeled the impacts of drought on vital rates explicitly and created general, density-independent, stochastic, parameter-resampled IPMs (*sensu* @metcalfStatisticalModellingAnnual2015). We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016] as described in @scottDelayedEffectsClimate2022. For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months [@scottDelayedEffectsClimate2022] (@fig-lifecycle). To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data. Then, 36 month lags are calculated for each year starting in February (the month of the demographic census). These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM. With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.\

<!--# ergodicity is somethign discussed in the Caswell book.  Basically, kernels eventually "forget" previous states.  I don't know if this last sentence really needs to be included, or if we are just inviting criticism from reviewers. -->

All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package in R [@levinIpmrFlexibleImplementation2021]. The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation? -->. For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects. Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and back-transformed to be on the same scale as deterministic lambdas for comparison. We used the distribution of established plant sizes and proportion of seedlings from the full dataset as a starting population vector. While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.\

To estimate uncertainty around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat. For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above. We then used these bootstraped estimates of lambda to calculate bias corrected 95% confidence intervals [@ellnerDatadrivenModellingStructured2016].\

This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.\

### *Statistical analyses*

> All about the stats.

## Results & Discussion

```{r }
meta_df <- 
  tar_meta(store = here("_targets")) %>% 
  select(name, size, bytes, time, seconds) %>% 
  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
  mutate(minutes = seconds / 60,
         hours = minutes / 60)

meta_df_tbl<-meta_df %>% 
  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
  select(-trash) %>% 
  select(IPM, Habitat, minutes) %>% 
  group_by(IPM) %>% 
mutate(IPM = str_replace_all(
    IPM,
    c(
      "det" = "det",  # Deterministic",
      "stoch" = "sk", # Stochastic, kernel-resampled
      "dlnm" = "sp" # Stochastic, parameter-resampled
    ))) %>%
  summarize(mean_time_min = round(mean(minutes), 2)) %>%
  rename("IPM Type" = IPM) %>% 
  rename("mean time (min.)" = mean_time_min)

det_time<-meta_df_tbl %>% filter(`IPM Type`=="det")
det_time<-as.numeric(det_time[1,2])
sker_time<-meta_df_tbl %>% filter(`IPM Type`=="sk")
sker_time<-as.numeric(sker_time[1,2])
spar_time<-meta_df_tbl %>% filter(`IPM Type`=="sp")
spar_time<-as.numeric(spar_time[1,2])
```

1.  For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (@tbl-aic).\
    <!--# No longer true if random effect = different smooth for each year -->

2.  Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (@tbl-lambdas).\

3.  We found that the the choice of Integral Projection Model didn't change the relative ranking of lambda in Continous Forest and Fragments.\

4.  DLNM models take much, much longer to iterate: while the Deterministic and Kernel-resampled stochastic models took \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Parameter-resampled stochastic models with lagged effects took \~`r spar_time` min.\

5.  The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.\

<!--# also note that this can't really be paralellized because each iteration depends on the previous one.  There's *maybe* some potential speed gains in using multiple cores with predict.bam(), but unlikely to be substantial. -->

6.  @fig-pop-states has some interesting things in it:

> -   For the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e. more medium sized plants in FF).

> -   For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF

> -   For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF. This makes sense as we know lagged effects are more important in fragments.

> -   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in @fig-pop-states (B)

> -   Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF

> -   These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.

## Acknowledgments

We thank **,** \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript. We thank Sam Levin for his help with the `ipmr` package. Financial support was provided by the U.S. National Science Foundation (awards \_\_\_\_, and \_\_\_\_). This article is publication no. -- -- in the BDFFP Technical series. The authors declare no conflicts of interest.

## CRediT Statement

<!--# https://casrai.org/credit/ -->

ERS contributed to the conceptualization, methodology, formal analysis, and led the writing of the original draft. EMB contributed to the conceptualization, methodology, and writing and also acquired funding.

## Data Availability Statement

Data and R code used in this study are archived with Zenodo at <doi url>.

<!-- ```{r } -->

<!-- pkgs <- cite_packages(output = "table", out.dir = ".") -->

<!-- knitr::kable(pkgs) -->

<!-- ``` -->

## Literature Cited

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\pagebreak

<!-- # Tables -->

```{r}
library(pander)
library(glue)
library(snakecase)
```

```{r aic, results='asis'}
#| label: tbl-aic
#| tbl-cap: "Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
#| 
#TODO: only show first instance of habitat and vital rate columns, not on every row.  Alternatively do something fancy with flextable or gt.

aic_tbl_df %>%
  group_by(habitat, model) %>% 
  # mutate(rank = row_number()) %>% 
  ungroup() %>% 
  mutate(habitat = toupper(habitat),
         env_effect_incl = ifelse(
           env_effect_incl != "DLNM",
           to_sentence_case(as.character(env_effect_incl)),
           as.character(env_effect_incl)
         )
  )%>% 
  mutate(
    model = fct_relevel(
      model,
      c("surv", "size", "flwr", "surv_sdlg", "size_sdlg")
    ),
    model = fct_recode(
      model,
      "Survival"          = "surv",
      "Growth"            = "size",
      "Flowering"         = "flwr",
      "Seedling survival" = "surv_sdlg",
      "Seedling growth"   = "size_sdlg"
    )
  ) %>% 

   #another option is to use psuedo model notation.  Can't fit both unless I remove the "habitat" column or something.
  mutate(RHS = str_replace_all(
    env_effect_incl,
    c(
      "None" = "~s(z)",
      "Random effect of year" = "~s(z, by = 'year')",
      "DLNM" = "~s(z) + te(spei, lag)"
    )
  )) %>% 
  
  arrange(habitat, model) %>% 
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "edf" = "df",
    "\u0394AIC" = "dAIC"
  ) %>% 
  
pandoc.table(
  justify = c( "left", "left","left", "right",  "right"),
  full_width = FALSE
  # ,
  # caption = Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
)
```

\newpage

```{r tbl-lambdas, results='asis'}
#| label: tbl-lambdas
#| tbl-cap: "Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."

lambda_table %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "DLNM" = "Stochastic, parameter-resampled"
    )
  )) %>% 
  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
pandoc.table(
  digits = 5,
  keep.trailing.zeros = TRUE,
  justify = c("left", "left", "right"),
  full_width = FALSE
  # ,
  # caption = "(\\#tab-lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
)

```

\newpage

<!-- # Figures -->

```{r #fig-lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Lifecycle diagram of *Heliconia acuminata*. Each transition is associated with an equation for a vital rate function.  The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM. The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."

# Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
path <- here("docs/figures/lifecycle.png")
knitr::include_graphics(path, rel_path = FALSE)
```

\newpage

```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}

#| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations. Stacked area charts (A) show the relative size/stage distribution of plants in continuous forest (CF, top row) and forest fragments (FF, bottom row) in each of the three IPMs (columns). The proportion of each size class in CF and FF for each iteration is shown in B with the first 30 iterations removed to not include transient dynamics. A 1:1 line is plotted in black.  Size categories include seedlings (a discrete category in the IPMs), pre-reproductive 1 (log(size) 0–2.5) that have low average survival (< 0.9) and a near 0 probability of flowering, pre-reproductive 2 (log(size) 2.5–4.5) that have a higher average survival probabilty (> 0.8) and a near 0 probability of flowering, reproductive 1 (log(size) 4.5–6) that have a high average survival probability (>0.95) and a lower flowering probability (< 0.25), and reproductive 2 (log(size) 6+) that have a high average survival probability (>0.95) and higher flowering proability (> 0.2)."

knitr::include_graphics(here(fig_pop_states), rel_path = FALSE)
```
