---
title: "Context-dependent consequences of lagged effects in demographic models"
shorttitle: "Lagged Effects IPMs"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affiliations:
      - ref: 1
    corresponding: true
  - name: María Uriarte
    email: mu2126@columbia.edu
    affiliations:
      - ref: 2
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affiliations:
      - ref: 1
      - ref: 3
      - ref: 4
affiliations:
  - id: "1"
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: "2" 
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: "3"
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: "4"
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format: 
  pdf:
    # template: AmNat_MS_template.tex
    # documentclass: scrarticle
    documentclass: article
    number-sections: false
    colorlinks: true
    keeptex: true
    fontsize: 12pt
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{sectsty}\sectionfont{\centering}
        \usepackage{sectsty}\subsectionfont{\centering}
        \usepackage{fontspec}
        \usepackage{titlesec}
        <!-- \usepackage[sc]{mathpazo} %Like Palatino with extensive math support -->
        <!-- \usepackage[utf8]{inputenc} -->
        \usepackage{parskip}
        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
        \setlength{\parindent}{0pt}
    
        
#\sectionfont{\color{black}\fontsize{12}{16.8}\selectfont}
#\subsectionfont{\color{black}\fontsize{12}{16.8}\selectfont}
        
# number after font size is the interline spacing. It should about 20-25 % more 
# than the font size (for a 10 pt document, you can measure it is equal to 12 pt).
# \titleformat{\subsection}
# \titleformat*{\subsection}{\fontsize{12{16.8}\selectfont
        #     {\sffamily\Large\bfseries}{\thesection}{1em}{}[{\titlerule[0.8pt]}]

# When ready to submit, these will change to double space and add a 
# paragraph indent
# \usepackage{setspace}\doublespacing
# \setlength\parindent{22pt}

fig-cap-location: bottom
tbl-cap-location: top
geometry:
  - top=1in
  - left=1in
  - bottom=1in
  - right=1in
keywords: 
  - integral projection models
  - demography
  - population dynamics
  - environmental stochasticity
  - lagged effects
  
Notes: >
  Manuscript elements: Figure~1, figure~2, table~1, appendix~A.  
  Manuscript type: e-note


# abstract: >
#  The abstract should have 150 words. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(grateful)
library(tidyverse)
library(kableExtra)
library(here)
```

```{r}
#load targets
withr::with_dir(here(), {
  tar_load(c(
    aic_tbl_df,
    lambda_table,
    fig_pop_states
  ))
})
```

<!-- The actual document text starts here: -->

\pagebreak

# Abstract

<!-- AmNat Note Abstracts are no longer than 150 words. -->

Text of 150 words max summarizing this amazing paper.

\bigskip

**Keywords:** demography, environmental stochasticity, integral projection models, lagged effects, structured population models, population dynamics

\bigskip

**Manuscript elements:** Figure\~1, figure\~2, table\~1, appendix\~A..

\bigskip

**Manuscript type:** e-note

\pagebreak

<!-- AmNat Notes: communicate concise points, using either data or theory. Like Major Articles, they present insights of broad general significance and interest. Notes, on average, should be no more than 3000 words of text (not including the Literature Cited) and have no more than three figures and/or tables in print. -->

## Introduction

There is increasing evidence that an organism's current likelihood of growth, survival, or reproduction can be strongly influenced by previous environmental conditions. These *Lagged Effects*, also known as *Delayed Life-history Events* (i.e., DLHEs) [@beckerman_population_2002], can simultaneously affect an entire cohort (e.g., juveniles hatching during a period of scarcity will all have delayed maturation and lower lifetime fecundity) or only a subset of the population (e.g., cold temperatures in one year lead to reduced flowering by potentially reproductive individuals in the next). In addition, the temporal delay between an environmental event and changes in demographic vital rates depends on both the intensity of the event and its timing relative to the underlying physiological processes [@eversLaggedDormantSeason2021; @criley_year_1994]. A drought during the early stages of gestation or floral bud formation, for example, might have a much larger impact on the number of fruits or offspring produced than one several months later whose timing coincides with birth or flowering. The delay or intensity of lagged effects can also depend on local ecological conditions, with individuals in some habitats buffered against -- or able to recover more quickly from - the delayed effects of environmental variation.\

Because Lagged Effects are often directly linked to reproduction and survival, it is thought they could have major consequences for population dynamics [@beckerman_population_2002]. Although there is emerging evidence that this is indeed the case [e.g., @molowny-horas_changes_2017; @tenhumberg_time-lagged_2018; @williams_life_2015], broader efforts to test this hypothesis have been hampered by two primary factors: First, detecting lagged effects requires long-term data on both the putative lagged effect (i.e., probability of flowering) and its potential environmental drivers [@metcalfStatisticalModellingAnnual2015]. These coupled data sets are rare [sensu @eversLaggedDormantSeason2021], in part because studies to disentangle lagged effects can be challenging to design and maintain [@kuss_evolutionary_2008]. Second, the methods for identifying lagged effects and modeling their demographic impacts can be challenging to implement. Many of the statistical methods have stringent data requirements [@metcalfStatisticalModellingAnnual2015] and assumptions, while the including complex biological processes in demographic models can render them less tractable. Addressing these obstacles is a major undertaking; the value of doing so will depend on the effort required vs. the potential consequences of failing to consider lagged effects - consequences that range from overestimating projections of population growth rate (i.e., $\lambda$) in a conservation setting to drawing invalid conclusions regarding support for the predictions of ecological or evolutionary theory.\

Integral Projection Models (i.e., IPMs) are an important and widely used tool for studying demography and population dynamics [@ellner_integral_2006; @rees_building_2014; @rees_integral_2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguerogomez_compadre_2015], have facilitated their use to study a wide range of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morris_quantitative_2002 and @crone_how_2011]. Mathematical and statistical advances [e.g., @williams2012avoiding; @brooks_statistical_2019] have rapidly expand the scope of questions and biological processes that can be investigated with these models [e.g., @metcalfStatisticalModellingAnnual2015; @rees_evolving_2016; @ellnerDatadrivenModellingStructured2016]. Here we investigate how including lagged effects in Integral Projection Models influences projections of $\lambda$ and population structure.\

We have previously shown that the effects of precipitation extremes on the demographic vital rates of an Amazonian understory herb (*Heliconia acuminata*, Heliconiaceae) can be delayed up to 36 months [@scottDelayedEffectsClimate2022], with the presence and duration of these lagged effects varying by vital rate and habitat. We parameterized three classes of Integral Projection Models - a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates - for populations in two habitat type (i.e., continuous forest vs. forest fragments). Based on previous studies [@bruna_are_2003; @bruna_demographic_2005; @bruna_experimental_2002; @bruna_habitat_2002] and demographic theory [@caswell2001; @tuljapurkar_population_1990] we predicted that:\

(i) projections of $\lambda$ would be lower in forest fragments than in continuous forest for all model types,
(ii) projections of $\lambda$ from deterministic models would be higher than those of stochastic models,
(iii) projections of $\lambda$ would be lowest for models with lagged effects, and
(iv) populations would be more skewed towards pre-reproductive size classes in fragments that forest, regardless of model structure

## Methods

### *Study System and Demographic Data*

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kress_diversity_1990]. While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others - including *H. acuminata* - grow primarily in the forest understory [@kress_self-incompatibility_1983; @ribeiro_influence_2010]. Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@bruna_heliconia_2004; @stouffer_forest_1996]. The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H. acuminata* found at Brazil's Biological Dynamics of Forest Fragments Project (BDFFP), located \~70 km north of Manaus, Brazil. The BDFFP reserves include both continuous forest and forest fragments that range in size from 1-100 ha. These fragment reserves were originally isolated in the early 1980’s by the creation of cattle pastures, with the secondary growth surrounding them periodically cleared to ensure their continued isolation. The habitat in all sites is non-flooded lowland rain forest with rugged topography. A complete summary of the BDFFP and its history can be found in @bierregaard_lessons_2001.\

A complete description of the demographic methods, data, and analyses to date can be found in @bruna_demography_2023. Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments. All of the *Heliconia acuminata* in these plots were marked and measured; the plots were censused annually, at which time a team recorded the size of surviving individuals, marked and measured new seedlings, and identified any previously marked plants that died. Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants; in our site *H. acuminata* begin flowering early in the rainy season (e.g., January) and most reproductive plants produce a single infloresence (range = 1--7) with 20--25 flowers [@bruna_habitat_2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarte_disentangling_2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@bruna_habitat_2002; @bruna_seed_1999]. On average olots in CF also had more than twice as many plants as the plots in 1-ha fragments (CF median = 788, range = (201-1549); 1-ha median = 339, range = (297-400)).\

### *Construction of Integral Projection Models*

We compared three classes of IPMs - Deterministic, Stochastic, and Stochastic with lagged effects of precipitation - each of which required different functional forms of their underlying vital rates models (@fig-lifecycle). All of the models were density-independent.

**(1) Deterministic IPM:** In this model, which served as the foundation for more complex models, the number and size of mature plants in year $t+1$ is determined by sub-kernels describing the survival and growth (or regression) of mature plants from year $t$ to $t+1$, the number of seedlings establishing in year $t$ that survive to $t+1$ (@eq-mature), and the number of new seedlings that enter the population (@eq-sdlg).

$$
n(z^{\prime},t+1) = R(z^\prime)n_s(t) + \int_L^U P(z^\prime,z) n(z,t)\;dz
$$ {#eq-mature}

$$
n_s(t+1) = \int_L^U F(z) n(z,t) \; dz
$$ {#eq-sdlg}

In @eq-mature the sub-kernel $P(z^\prime, z)$ describes the size-dependent survival and growth or regression ($s(z)$ and $G(z^\prime, z)$, respectively) of mature plants from one year to the next ([@eq-P]). Sub-kernel $R(z^\prime)$ describes the survival and growth of seedlings that established in year $t$ to year $t+1$ ([@eq-R]); note that these values are not size-dependent - in all our IPMs 'seedlings' are a discrete, size-independent category with a probability $s_s$ of survival for one year following germination and establishment, with the size at which they transition into the 'mature' plant category given by $G(z^\prime, z)$. IPMs can include transitions between continuous and discrete states [@ellnerDatadrivenModellingStructured2016]); we treat seedlings as a distinct category because they have lower survival and growth in their first year than comparably sized plants [@bruna_are_2003; @scottDelayedEffectsClimate2022].

$$
P(z^\prime, z) = s(z)G(z^\prime, z)
$$ {#eq-P}

$$
R(z^\prime) = s_sG_s(z^\prime)
$$ {#eq-R}

The number of new seedlings entering the population, $n_s(t+1)$ ([@eq-sdlg]), is described by the fecundity sub-kernel ([@eq-F]). Both the probability that a mature plant will flower, $p_f(z)$, and the number of seeds they produce, $f(z)$, are size dependent. These seeds germinate and establish as seedlings with probability $g$.

$$
F(z) = p_f(z)f(z)g
$$ {#eq-F}

Vital rate models for growth ($G_s(z^\prime)$ and $G(z^\prime, z)$), survival ($s_s$ and $s(z)$), and flowering ($p_f(z)$) were fit using the long-term demographic data [@bruna_are_2003]. For established plants, these three vital rates were modeled as a smooth function of size in the previous census using generalized additive models (GAMs) fit with the `mgcv` package [@mgcv] in `r version$version.string` [@rcoreteam2020]<!--# probably include a supplement with detail on things like basis, number of knots, etc. -->. For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. intercept only models). For growth models ($G_s(z^\prime)$ and $G(z^\prime, z)$) a scaled t family distribution provided a better fit to the data than a gaussian fit as the residuals were leptokurtic with a simple Gaussian model. Size-specific fecundity was based on surveys of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] (together $f(z)$) and experimentally derived estimates of germination and establishment in forest and fragments [@bruna_effects_2002; @bruna_seed_1999] <!--# Emilio: add more info here -->.\

**(2) Stochastic IPM:** To build the stochastic, kernel-resampled IPMs we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (@fig-lifecycle). The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years. The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly. This is equivalent to the matrix selection approach for matrix population models described by @caswell2001.\

**(3) Stochastic IPM with lagged effects of precipitation on vital rates:** In our The stochastic, parameter-resampled IPMs the impacts of precipitation extremes on vital rates were modeled explicitly (*sensu* @metcalfStatisticalModellingAnnual2015). We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016] as described in @scottDelayedEffectsClimate2022. For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months [@scottDelayedEffectsClimate2022] (@fig-lifecycle). To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data. Then, 36 month lags are calculated for each year starting in February (the month of the demographic census). These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM. With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.\

<!--# ergodicity is somethign discussed in the Caswell book.  Basically, kernels eventually "forget" previous states.  I don't know if this last sentence really needs to be included, or if we are just inviting criticism from reviewers. -->

All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package in R [@levinIpmrFlexibleImplementation2021]. The IPMs used 100 meshpoints and the midpoint rule for calculating kernels <!--# citation? -->. For each type of IPM we iterated the model for 1000 time steps, discarding the first 100 time steps to omit transient effects. Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswell2001] and back-transformed to be on the same scale as deterministic lambdas for comparison. We used the distribution of established plant sizes and proportion of seedlings from the full dataset as a starting population vector. While other starting population vectors were possible, the choice is of little importance as it will only impact transient dynamics, which we aren't interested in for this study.\

To estimate uncertainty around the per-capita growth rates (lambdas), we created 500 bootstraps of the demographic dataset by sampling individual plants with replacement within each habitat. For each bootstrap, we then re-fit vital rates models (all except germination and establishment rate, fruits per flowering plant, and seeds per fruit, which were estimated using different datasets), constructed IPMs, and calculated a value for lambda as described above. We then used these bootstraped estimates of lambda to calculate bias corrected 95% confidence intervals [@ellnerDatadrivenModellingStructured2016].\

#TODO: need to add something about the comparison of population structure (figure)

This workflow was managed using the `targets` R package [@targets] which also allowed us to track computational time spent on each IPM for comparison.\

### *Statistical analyses*

Comparison of CF vs FF

-   For det need to bootstrap

-   for within stoch and lag: glm/t-test.

## Results & Discussion

```{r }
meta_df <- 
  tar_meta(store = here("_targets")) %>% 
  select(name, size, bytes, time, seconds) %>% 
  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
  mutate(minutes = seconds / 60,
         hours = minutes / 60)

meta_df_tbl<-meta_df %>% 
  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
  select(-trash) %>% 
  select(IPM, Habitat, minutes) %>% 
  group_by(IPM) %>% 
mutate(IPM = str_replace_all(
    IPM,
    c(
      "det" = "det",  # Deterministic",
      "stoch" = "sk", # Stochastic, kernel-resampled
      "dlnm" = "sp" # Stochastic, parameter-resampled
    ))) %>%
  summarize(mean_time_min = round(mean(minutes), 2)) %>%
  rename("IPM Type" = IPM) %>% 
  rename("mean time (min.)" = mean_time_min)

det_time<-meta_df_tbl %>% filter(`IPM Type`=="det")
det_time<-as.numeric(det_time[1,2])
sker_time<-meta_df_tbl %>% filter(`IPM Type`=="sk")
sker_time<-as.numeric(sker_time[1,2])
spar_time<-meta_df_tbl %>% filter(`IPM Type`=="sp")
spar_time<-as.numeric(spar_time[1,2])
```

1.  For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (@tbl-aic).\
    <!--# No longer true if random effect = different smooth for each year -->

2.  Population growth rates in CF were significantly higher than in forest fragments for all IPM types (@tbl-lambdas).\

3.  Deterministic and stocuhastic-kernal similar means.

4.  These results are consistent with those of @kayeEffectStochasticTechnique2003, who found who found that the method effected stochastic lambda but relative ranking of populations was consistent.

5.  BUT lag dropped lambdas considerably. In fact, took both well below zero. This is a big, big difference. If you are doing conservation and management, a growth rate lower by 5-6 % would really rock your world.

### Population Structure

6.  @fig-pop-states has some interesting things in it:

> -   For the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e. more medium sized plants in FF).

> -   For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF

> -   For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF. This makes sense as we know lagged effects are more important in fragments.

> -   Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in @fig-pop-states (B)

> -   Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF

### Time and Effort

4.  DLNM models take much, much longer to iterate: while the Deterministic and Kernel-resampled stochastic models took \~`r det_time` and \~`r sker_time` min to iterate (respectively), the Parameter-resampled stochastic models with lagged effects took \~`r spar_time` min.\

5.  The greater use of computational resources is likely a result of `predict()` being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.\

<!--# also note that this can't really be paralellized because each iteration depends on the previous one.  There's *maybe* some potential speed gains in using multiple cores with predict.bam(), but unlikely to be substantial. -->

## Acknowledgments

We thank **,** \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript. We thank Sam Levin for his help with the `ipmr` package. Financial support was provided by the U.S. National Science Foundation (awards \_\_\_\_, and \_\_\_\_). This article is publication no. -- -- in the BDFFP Technical series. The authors declare no conflicts of interest.

## CRediT Statement

<!--# https://casrai.org/credit/ -->

ERS contributed to the conceptualization, methodology, formal analysis, and led the writing of the original draft. EMB contributed to the conceptualization, methodology, and writing and also acquired funding.

## Data Availability Statement

Data and R code used in this study are archived with Zenodo at *(doi and url to be added on acceptance)*.

<!-- ```{r } -->

<!-- pkgs <- cite_packages(output = "table", out.dir = ".") -->

<!-- knitr::kable(pkgs) -->

<!-- ``` -->

\newpage

## Literature Cited

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\pagebreak

<!-- # Tables -->

```{r}
library(pander)
library(glue)
library(snakecase)
```

```{r aic, results='asis'}
#| label: tbl-aic
#| tbl-cap: "Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
#| 
#TODO: only show first instance of habitat and vital rate columns, not on every row.  Alternatively do something fancy with flextable or gt.

aic_tbl_df %>%
  group_by(habitat, model) %>% 
  # mutate(rank = row_number()) %>% 
  ungroup() %>% 
  mutate(habitat = toupper(habitat),
         env_effect_incl = ifelse(
           env_effect_incl != "DLNM",
           to_sentence_case(as.character(env_effect_incl)),
           as.character(env_effect_incl)
         )
  )%>% 
  mutate(
    model = fct_relevel(
      model,
      c("surv", "size", "flwr", "surv_sdlg", "size_sdlg")
    ),
    model = fct_recode(
      model,
      "Survival"          = "surv",
      "Growth"            = "size",
      "Flowering"         = "flwr",
      "Seedling survival" = "surv_sdlg",
      "Seedling growth"   = "size_sdlg"
    )
  ) %>% 

   #another option is to use psuedo model notation.  Can't fit both unless I remove the "habitat" column or something.
  mutate(RHS = str_replace_all(
    env_effect_incl,
    c(
      "None" = "~s(z)",
      "Random effect of year" = "~s(z, by = 'year')",
      "DLNM" = "~s(z) + te(spei, lag)"
    )
  )) %>% 
  
  arrange(habitat, model) %>% 
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "edf" = "df",
    "\u0394AIC" = "dAIC"
  ) %>% 
  
pandoc.table(
  justify = c( "left", "left","left", "right",  "right"),
  full_width = FALSE
  # ,
  # caption = Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
)
```

\newpage

```{r tbl-lambdas, results='asis'}
#| label: tbl-lambdas
#| tbl-cap: "Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."

lambda_table %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic, kernel-resampled",
      "DLNM" = "Stochastic, parameter-resampled"
    )
  )) %>% 
  select(IPM = ipm, Habitat = habitat, "$\\lambda$" = lambda) %>% 
pandoc.table(
  digits = 5,
  keep.trailing.zeros = TRUE,
  justify = c("left", "left", "right"),
  full_width = FALSE
  # ,
  # caption = "(\\#tab-lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
)

```

\newpage

<!-- # Figures -->

```{r #fig-lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Lifecycle diagram of *Heliconia acuminata*. Each transition is associated with an equation for a vital rate function.  The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM. The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."

# Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
path <- here("docs/figures/lifecycle.png")
knitr::include_graphics(path, rel_path = FALSE)
```

\newpage

```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}

#| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations. Stacked area charts (A) show the relative size/stage distribution of plants in continuous forest (CF, top row) and forest fragments (FF, bottom row) in each of the three IPMs (columns). The proportion of each size class in CF and FF for each iteration is shown in B with the first 30 iterations removed to not include transient dynamics. A 1:1 line is plotted in black.  Size categories include seedlings (a discrete category in the IPMs), pre-reproductive 1 (log(size) 0–2.5) that have low average survival (< 0.9) and a near 0 probability of flowering, pre-reproductive 2 (log(size) 2.5–4.5) that have a higher average survival probabilty (> 0.8) and a near 0 probability of flowering, reproductive 1 (log(size) 4.5–6) that have a high average survival probability (>0.95) and a lower flowering probability (< 0.25), and reproductive 2 (log(size) 6+) that have a high average survival probability (>0.95) and higher flowering proability (> 0.2)."

knitr::include_graphics(here(fig_pop_states), rel_path = FALSE)
```
