---
title: Context-dependent consequences of including delayed responses to climate in Integral Projection Models 

shorttitle: "IPMs with Lagged Effects"  
authors:
  - name: Eric R. Scott
    email: scottericr@gmail.com
    affil-id: 1,5
    orcid: 0000-0002-7430-7879
  - name: María Uriarte
    email: mu2126@columbia.edu
    affil-id: 2
    orcid: 0000-0002-0484-0758
  - name: Emilio M. Bruna
    email: embruna@ufl.edu
    affil-id: 1,3,4
    corresponding: true
    orcid: 0000-0003-3381-8477
affiliations:
  - id: 1
    name: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430 USA
  - id: 2
    name: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - id: 3
    name: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530 USA
  - id: 4
    name: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
  - id: 5
    name: \textit{Current Address:} Communication & Cyber Technologies, University of Arizona, Tucson, AZ 85721 USA
date: 'last-modified'
date-format: '([draft:] D MMMM YYYY)'
linkcolor: blue
bibliography: lagged-ipms-ms.bib
csl: "templates/journal-of-ecology.csl" # Insert path for the bib-style
# 
# highlights: |
#   These are the highlights
format:
  # docx:
  #   prefer-html: true
  pdf:
    # template: AmNat_MS_template.tex
    # documentclass: scrarticle
    # documentclass: article
    number-sections: true
    # colorlinks: true
    # keeptex: true
    fontsize: 12pt
    # fig-pos: "H"
    # the following line is to render the aaffiliations in pdf, which is not default in quarto, see 
    # https://stackoverflow.com/questions/75040607/why-do-affiliations-not-show-up-anywhere-in-the-pdf-output-of-quarto
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\normalsize}
        \usepackage{lineno}\linenumbers
        \usepackage{fontspec}
        \usepackage{titlesec}
        \usepackage{parskip}
        \setlength{\parskip}{1pt} % 1ex plus 0.5ex minus 0.2ex}
        \usepackage{caption}
        \captionsetup[Figure]{labelfont={bf},name={FIGURE},labelsep=period} 
        \raggedright
        \usepackage{setspace}\doublespacing
        \setlength\parindent{22pt} 
        <!-- # for heading sizes and spacing below: -->
        <!-- # number after font size is the interline spacing. It should about 20-25 % more  -->
        <!-- # than the font size (for a 10 pt document, you can measure it is equal to 12 pt). -->
        \usepackage{sectsty}\sectionfont{\color{black}\fontsize{15}{12}\bfseries} 
        \usepackage{sectsty}\subsectionfont{\normalfont\color{black}\fontsize{14}{12}} 
        \usepackage{sectsty}\subsubsectionfont{\color{black}\fontsize{13}{12}\bfseries} 
        <!-- \usepackage{sectsty}\sectionfont{\centering} -->
        <!-- \usepackage{sectsty}\subsectionfont{\centering} -->
        <!-- \titleformat{\subsection} -->
        <!-- \titleformat*{\subsection}{\fontsize{12{16.8}\selectfont -->
        <!--      {\sffamily\Large\bfseries}{\thesection}{1em}{}[{\titlerule[0.8pt]}] -->

fig-cap-location: bottom
tbl-cap-location: top
geometry:
  - top=1in
  - left=1in
  - bottom=1in
  - right=1in
keywords: 
  - Amazonia
  - habitat fragmentation
  - Heliconia 
  - lambda
  - plant demography
  - population dynamics
  - structured population models
  - vital rates

# filters:
#   - wordcount
# wordcount:
#   count-code-blocks: false
#   count-inline-code: false
  
Notes: >
  Manuscript elements: Figures 1-4, Tables 1-6, Supplementary Tables 1-2
  
  
  


# abstract: >
#  The abstract should have 150 words. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(grateful)
library(tidyverse)
library(kableExtra)
library(here)
# library(vartest)
library(purrr)
library(broom)
```

```{r}
#load targets
withr::with_dir(here(), {
  tar_load(c(
    aic_tbl_df,
    lambda_table,
    fig_pop_states
  ))
})

lambda_vars<-read_csv(here("docs","figures","lambda_vars.csv"))
stats_table_lambdavars<-read_csv(here("docs","figures","stats_table_lambdavars.csv"))
```

\newpage

# Abstract {.unnumbered}

1.  There is increasing evidence that changes in demographic vital rates due to contemporary climatic conditions can be delayed months or even years. This temporal delay between an climatic event and the resulting changes in vital rates can depend on both the intensity of the event and its timing relative to the physiological mechanisms underlying the response.
2.  Because lagged effects are often directly linked to reproduction and survival, it has been argued they could have major but underappreciated consequences for population dynamics. There have been few tests of this hypothesis, however, in part because the long-term data required to incorporate lagged effects in structured population models are rarely available.
3.  We parameterized three classes of Integral Projection Models --- deterministic, stochastic, and stochastic with lagged effects of precipitation on vital rates --- using a decade of annual censuses from populations of an Amazonian understory herb. We then evaluated whether (1) if incorporating lagged effects altered projections of population growth rate, and (2) if habitat-specific differences in lagged effects translate into habitat-specific projections of population growth rate (i.e., $\lambda$).
4.  Projections of $\lambda$ were lower when models included lagged effects. Projections of $\lambda$ and population structure were also more variable, though the magnitude of this variation varied by habitat type.\
5.  Synthesis: The choice of demographic model, and whether that model includes the lagged effects of climate, may be especially critical in studies with implications on endangered species or when attempting to identify populations that are declining. This is because failing to include lagged effects may overestimate the actual growth rate of a focal population. It can also lead to incorrect conclusions regarding the outcome of alternative management strategies, because even small differences in $\lambda$ can have major consequences for population dynamics.

<!-- \bigskip -->

<!-- \noindent{}**Keywords:** Amazonia, habitat fragmentation, *Heliconia*, lambda, plant demography, population dynamics, structured population models, vital rates -->

<!-- \bigskip -->

<!-- \noindent{}**Manuscript elements:** Figures 1-4, Tables 1-4, Supplementary Tables S1-S2 -->

<!-- \noindent{}**Manuscript type:** Article -->

\newpage

# Introduction

Current climatic conditions can have immediate impacts on the growth, survival, or reproduction of plants [@eversInclusionImmediateLagged2023; @williamsLinkingClimateDemography2025]. However, there is now ample evidence for delays of months or even years between current conditions and any resulting changes in demographic vital rates [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2021; @pratherCurrentLaggedClimate2023]. These *Lagged Effects*, also known as *Delayed Life-history Events* [@beckermanPopulationDynamicConsequences2002] or *Temporally Varied Responses* [@eversInclusionImmediateLagged2023], can simultaneously affect an entire cohort (e.g., all seedlings emerging during a drought greater risk of mortality due to pathogens) or only a subset of the population (e.g., a freeze this year reduces the probability of flowering, but only for trees below a particular size threshold). Moreover, the delay between an environmental event and altered vital rates can depend on both the intensity of the event and its timing relative to the underlying physiological mechanisms [@eversLaggedDormantSeason2021; @crileyYearProductionHigh1994]. For example, a drought during the early phase of floral bud formation might have a much larger impact on the number of flowers of fruits a plant produces than one at a later stage of development [sensu @tenhumbergTimelaggedEffectsWeather2018].

Because many Lagged Effects appear to influence plant reproduction and survival [e.g., @pratherCurrentLaggedClimate2023], it has been argued they could have major consequences for population dynamics [@beckermanPopulationDynamicConsequences2002; @tenhumbergTimelaggedEffectsWeather2018]. Despite this, very few studies have explicitly tested for lagged effects of climate on vital rates [reviewed in @eversLaggedDormantSeason2021] or included lagged effects in demographic models [*sensu* @tenhumbergTimelaggedEffectsWeather2018; @kissUnderstandingEffectsWeather2024; @mensahDelayedEffectsWaterlimiting2025]. In one of the most comprehensive to do so, @eversInclusionImmediateLagged2023 found that the length of the lag (i.e., the "temporal window") influenced the effect of climate variability on long-term population growth rates (i.e., $\lambda$; see also @tenhumbergTimelaggedEffectsWeather2018). However, it remains unknown whether models with and without lagged effects actually project different values of $\lambda$ or how the effects of temporal lags compare with those of other demographic drivers. Also unknown is the extent to which the demographic consequences of lagged effects vary among or between populations [@mensahDelayedEffectsWaterlimiting2025; @scottDelayedEffectsClimate2021]; if individuals in some locations or (micro)habitats are buffered against climatic extremes [@williamsLinkingClimateDemography2025], this could alter the magnitude of delayed effects or which vital rates they affect [@scottDelayedEffectsClimate2021]. A general understanding of how climate shapes plant demography requires understanding not only when and how to incorporate lagged effects in models [@eversInclusionImmediateLagged2023], but also how the demographic impacts of climate compare to those of landscape structure, disturbance, biotic interactions, or other drivers of population dynamics [e.g., @hartmannLongtermDataIndicates2023].

Efforts to include lagged effects in structured population models are rare in part because experimental tests of these effects are challenging to design, implement, and maintain [@kussEvolutionaryDemographyLonglived2008]. While observational data can also be used to detect lagged effects, doing so requires long-term data on both the focal species' demography and putative climate drivers. Not only are such coupled data sets extremely rare [*sensu* @eversLaggedDormantSeason2021], the methods for both identifying lagged effects [e.g., @ogleQuantifyingEcologicalMemory2015a; @eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2021] and modeling their demographic impacts [@eversInclusionImmediateLagged2023; @mensahDelayedEffectsWaterlimiting2025] can be challenging to implement. For example, many of the potentially applicable statistical methods have stringent assumptions and data requirements rarely met by ecological data [@metcalfStatisticalModellingAnnual2015; @eversInclusionImmediateLagged2023]. Terms for complex biological processes such as lagged effects can also result in models for which mathematical solutions are intractable or analysis via simulation is more difficult. Addressing these challenges is a major undertaking, the value of which depends on the effort required to consider lagged effects with more complex models vs. the potential consequences of failing to include them --- consequences that range from overestimating projections of $\lambda$ for endangered species to drawing incorrect conclusions in tests of theory.

Integral Projection Models (i.e., IPMs) are an important tool for studying demography and population dynamics [@ellnerIntegralProjectionModels2006; @reesBuildingIntegralProjection2014; @reesIntegralProjectionModels2009]. Their flexibility, in concert with a rapidly growing suite of software, data, and other resources [@ellnerDatadrivenModellingStructured2016; @levinIpmrFlexibleImplementation2021; @salguero-gomezCOMPADREPlantMatrix2015], have accelerated their use to study a broad array of topics in ecology, evolution, and conservation [@ellnerDatadrivenModellingStructured2016; @morrisQuantitativeConservationBiology2002; @croneHowPlantEcologists2011] including how climate influences the demography and population dynamics of plants [@czachuraDemographicBackcastingReveals2020; @williamsLifeHistoryEvolution2015]. However, efforts to develop and test IPMs with lagged climatic effects remain limited, as most work to date has been conducted with matrix models [e.g., @eversInclusionImmediateLagged2023; @mensahDelayedEffectsWaterlimiting2025].

We sought to address this deficit using three classes of Integral Projection Models --- a deterministic IPM, a stochastic IPM, and a stochastic IPM with lagged effects of precipitation on vital rates. We parameterized these models with long-term demographic data collected from populations of the Amazonian understory herb *Heliconia acuminata* found in two distinct habitat types [@brunaDemographyUnderstoryHerb2023]. The lagged effects of precipitation extremes on the growth, reproduction, and survival *H. acuminata* can be delayed up to 36 months, with the duration and magnitude of these lagged effects varying by both vital rate and habitat type [@scottDelayedEffectsClimate2021]. That makes this system uniquely suited to address two questions fundamental to the study of climate and population dynamics: First, how different are the projections of $\lambda$ and population structure from IPMs with and without lagged effects of climate? Second, do habitat-specific differences in the lagged effects of climate on vital rates translate into distinct projections of $\lambda$?

<!-- from demographic theory [@caswellMatrixPopulationModels2001; @tuljapurkarPopulationDynamicsVariable1990]: (i) projections of $\lambda$ from deterministic models would be higher than those of stochastic models, and that (ii) $\lambda$ is lower for Forest Fragment than Continuous Forest populations regardless of model, but the difference between the two habitats is greatest for the IPM with lagged effects.  <!--- [MU: To me, these predictions come out of nowhere. A few sentences above to justify them would be helpful for the non-IPM reader] --->

# Materials and Methods

## Study System and Demographic Data

*Heliconia acuminata* (Heliconiaceae) is a perennial, self-incompatible monocot that is distributed throughout much of the Amazon basin [@kressDiversityDistributionHeliconia1990]. While some *Heliconia* species grow in large aggregations on roadsides, gaps, and in other disturbed habitats, others, including *H. acuminata*, grow primarily in the forest understory [@kressSelfincompatibilitySystemsCentral1983; @ribeiroInfluencePostclearingTreatment2010]. Understory *Heliconia* species produce fewer flowers and are pollinated by traplining hummingbirds [@brunaHeliconiaAcuminataReproductive2004; @stoufferForestFragmentationSeasonal1996]. The models and analyses here are based on 11 years (1998-2009) of demographic data collected on \>8500 *H. acuminata* found at Brazil's Biological Dynamics of Forest Fragments Project (BDFFP), located \~70 km north of Manaus, Brazil. The BDFFP reserves include both continuous forest and forest fragments that range in size from 1-100 ha. These fragment reserves were originally isolated in the early 1980’s by the creation of cattle pastures, with the secondary growth surrounding them periodically cleared to ensure their continued isolation. The habitat in all sites is non-flooded lowland rain forest with rugged topography. A complete summary of the BDFFP and its history can be found in @bierregaardLessonsAmazoniaEcology2001.\

A complete description of the demographic methods and data can be found in @brunaDemographyUnderstoryHerb2023. Briefly, in 1997--1998 a series of 5000 m^2^ plots were established in the BDFFP reserves: N=6 in Continuous Forest and N=4 in 1-ha Forest Fragments (i.e., CF and FF, respectively). All of the *Heliconia acuminata* in these plots were marked and measured; the plots were censused annually, at which time a team recorded the size of surviving individuals, marked and measured new seedlings, and identified any previously marked plants that died. Each plot was also surveyed 4-5 times during the flowering season to identify reproductive plants. These surveys were complemented by data on the number of fruits per flowering plant [@brunaLeafNumberLeaf2021] and seeds per fruit [@brunaDatasetHeliconiaAcuminata2014] that were collected outside of the demographic plots to avoid altering within-plot recruitment. Experiments were used to quantify the probability of seed germination and seedling establishment in both fragments and continuous forest [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999].

The rainy season in our sites is typically from late December through late May. *Heliconia acuminata* in our site begin flowering early in the rainy season and most reproductive plants produce a single infloresence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and have 1--3 seeds per fruit ($\bar{x}=2$) that are dispersed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate approximately 6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. The median number of plants in CF plots was over twice that of the plots in 1-ha fragments (CF: median = 788, range = 201-1549; 1-ha: median = 339, range = 297-400, @brunaDemographyUnderstoryHerb2023).

## Construction of Integral Projection Models

We projected the growth rate and structure of *Heliconia acuminata* with three classes of IPMs: (1) deterministic, (2) stochastic, and (3) stochastic with lagged environmental effects. Each of these IPMs required different functional forms of the underlying vital rate functions used to describe the *H. acuminata* life cycle (@fig-lifecycle). All models were density-independent, with the deterministic model serving as the foundation for the more complex models. Our modeling workflows were managed using the `targets` R package [@landauTargetsPackageDynamic2021]; in addition to ensuring reproducibility this allowed us to track computational time spent processing and analyzing each class of IPM.\

### Deterministic IPM

In this model the size and structure of a population in year $t+1$ is determined by the probability of survival and change in size of the $n$ plants of size $z$ that are alive in year $t$, given by

$$
n(z^{\prime},t+1) = \int_L^U P(z^\prime,z) n(z,t)\;\textrm{d}z + R(z^\prime)n_s(t)
$$ {#eq-mature}

\noindent{}plus the number of new seedlings that entered the population, given by

$$
n_s(t+1) = \int_L^U F(z) n(z,t) \; \textrm{d}z
$$ {#eq-sdlg}

<!-- The number and size of mature plants in year $t+1$ is determined by the survival and growth (or regression) of mature plants from year $t$ to $t+1$, the number of seedlings establishing in year $t$ that survive to $t+1$ (@eq-mature), and the number of new seedlings that enter the population (@eq-sdlg).\ -->

\noindent{}@eq-mature has two components. The first is sub-kernel $P(z^\prime, z)$, which describes the size-dependent probability of survival of post-seedling plants, $s$, and their growth (or regression) from size $z$ to $z^\prime$: $$
P(z^\prime, z) = s(z)G(z^\prime, z)
$$ {#eq-P}

\noindent{The} second component is sub-kernel $R(z^\prime)$. This sub-kernel describes the survival of seedlings to year $t+1$ of seedlings that established in year $t$ and the size of these surviving seedlings when they transition into the 'mature' plant population after surviving one year: $$
R(z^\prime) = s_sG_s(z^\prime)
$$ {#eq-R}

\noindent{Note} that in @eq-R both the probability that newly established seedlings survive their first year, $s_s$, and the size to which they grow at the end of this year, $G_s$, are size-independent. IPMs can readily accommodate such transitions of individuals from a discrete state to a continuous one [i.e., from 'seedling' to 'post-seedling' plant of size $z'$', @ellnerDatadrivenModellingStructured2016]; we treated seedlings as a distinct and discrete category in our models because they have lower survival and growth in their first year than comparably sized but plants that have been established for over one year [@brunaArePlantPopulations2003; @scottDelayedEffectsClimate2021].

The number of new seedlings entering the population in year $t+1$ is a function of the number of potentially reproductive plants in year $t$ and a sub-kernel describing the fecundity of these individuals: $$
F(z) = p_f(z)f(z)g
$$ {#eq-F}

\noindent{In} this sub-kernel both the probability that a mature plant will flower, $p_f$, and the number of seeds a flowering plant will produce, $f$, are size-dependent, but all seeds germinate and establish as seedlings with probability $g$.\

We used the annual census data [@brunaDemographyUnderstoryHerb2023] to fit the deterministic vital rate functions for growth, survival, and flowering in each habitat type (@fig-lifecycle). The data from all plots within a habitat class were pooled to create a single 'summary population' representing that habitat type; the summary populations were then used to fit the vital rate functions. A summary population is a superior means of synthesizing the demography of multiple populations than calculating an average vital rate for multiple populations because it corrects for the disproportionate weight that the low plant numbers in some size classes in some locations can give to vital rates [@horvitzSpatiotemporalVariationDemographic1995; @brunaArePlantPopulations2003]. For established plants the vital rates were modeled as a smooth function of size in the previous census with generalized additive models (GAMs) fit with the `mgcv` package [@mgcv] for the R statistical programming language [@rcoreteam2020].<!--`r version$version.string`---> <!--# probably include a supplement with detail on things like basis, number of knots, etc. --> For consistency, seedling survival and growth were also modeled using GAMs, but without size in the previous census as a predictor (i.e. 'intercept-only' models). For growth models a scaled t family distribution provided a better fit to the data than a Gaussian fit, as the residuals with a simple Gaussian model were leptokurtic. To model size-specific fecundity we used data from prior studies on the number of fruits per flowering plant [@brunaLeafNumberLeaf2021], seeds per fruit [@brunaDatasetHeliconiaAcuminata2014], and experimentally derived estimates of seed germination and seedling establishment [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999]. The plant size and stage categories to which plants were assigned, with their respective probabilities of survival and flowering, can be found in @tbl-plantcats.

### Stochastic IPMs

To include temporal stochasticity in our IPMs we included a random effect of year. This was done using a factor--smooth interaction that allowed the functional form of the relationship between plant size and vital rates to vary among transition years (@fig-lifecycle). We generated kernels for every transition year using the long-term survey data [@brunaDemographyUnderstoryHerb2023] and random smooths for year. We then randomly selected one of these sets of kernels to use in each iteration of the IPM. This procedure is equivalent to *'kernel resampling'* [*sensu* @metcalfStatisticalModellingAnnual2015] or matrix selection for matrix population models [@caswellMatrixPopulationModels2001; @boyceDemographyIncreasinglyVariable2006].

### Stochastic IPMs with lagged effects of precipitation on vital rates

We explicitly modeled the lagged effects of precipitation extremes on vital rates using the procedure described in @scottDelayedEffectsClimate2021. Briefly, we first calculated the Standardized Precipitation Evapotranspiraton Index (i.e., SPEI) for our study site using a published gridded dataset based on ground measurements [@xavierDailyGriddedMeteorological2016]. After we fit vital rate models using the long-term survey data (@fig-lifecycle), we modeled delayed effects of SPEI with Distributed Lag Non-linear Models (i.e., DLNMs) with a maximum lag of 36 months [@scottDelayedEffectsClimate2021]. To iterate these parameter-resampled IPMs [*sensu* @metcalfStatisticalModellingAnnual2015] we first created a random sequence of SPEI values by sampling years of the observed (monthly) SPEI data. For every year we then calculated a lag of 36 months from the month in which that year's census was completed. These values were then used to predict the fitted values from vital rate models, which generated different kernels for each iteration of the IPM. The kernels of successive iterations are not entirely independent --- the SPEI values used to calculate vital rates include values used in the previous two iterations --- but they are ergodic.\

<!--# ergodicity is somethign discussed in the Caswell book.  Basically, kernels eventually "forget" previous states.  I don't know if this last sentence really needs to be included, or if we are just inviting criticism from reviewers. -->

All IPMs were constructed and iterated <!--# optimized in notes/optimizing-meshpoints.Rmd which could be turned into supplemental -->using the `ipmr` package [@levinIpmrFlexibleImplementation2021] for the R statistical programming language [@rcoreteam2020]. The IPMs used 100 meshpoints and the midpoint rule for calculating kernels [@reesBuildingIntegralProjection2014]. For each type of IPM we iterated the model for 1000 time steps, but discarded the first 100 time steps to omit transient effects. Stochastic growth rates ($\lambda_s$) were calculated as the average $ln(\lambda)$ from each time step [@caswellMatrixPopulationModels2001] and then back-transformed to allow for direct comparison with projections of $\lambda$ from deterministic models. The initial starting vector primarily influences a population's transient dynamics; we therefore used the distribution of established plant sizes and proportion of seedlings in the full demographic data set as the initial population vector for all models.

Finally, we estimated the 95% confidence intervals for each IPM's projection of $\lambda$. To do so we first created 500 populations for each habitat type by sampling individual plants with replacement (i.e., bootstrapping) until the population size of each matched that of the initial population vector. We then re-fit the vital rate models for growth, survival, and flowering for each of these bootstrapped population and constructed new IPMs for each population as above (the models for germination and establishment rate, fruits per flowering plant, and seeds per fruit were not refit because these were estimated using different data sets). The projections of $\lambda$ for the new populations were then used to estimate the upper and lower 95% bias-corrected percentile intervals [@manlyRandomizationBootstrapMonte2018; @caswellMatrixPopulationModels2001].

## Statistical analyses

We used a Generalized Linear Model to compare the projections of $\lambda$ from the stochastic IPMs with and without lagged effects. We modeled $\lambda_{i}$ as a function of IPM type (i.e., with and without lagged effects) and habitat (i.e., continuous forest vs. fragment), and the interaction of IPM type and habitat. Because the response variable ($\lambda$) was continuous we used the Gaussian distribution with the identity link function in our model. We verified the model assumptions by plotting residuals versus fitted values.\
<!--- [MU: The use of a Gaussian GLM for λ comparisons is reasonable, but the response variable is bounded and potentially heteroscedastic. Have you considered a robust regression or bootstrapped ANOVA for λ differences?] --->

To determine if the deterministic projections of $\lambda$ for the populations in Continuous Forest (i.e, $\lambda^{CF}$) and Forest Fragments (i.e., $\lambda^{FF}$) were significantly different we used the randomization test procedure described by @caswellMatrixPopulationModels2001. Briefly, we randomly assigned plants with their demographic history among two populations, $R_{1}$ and $R_{2}$, that were equal in size to the original CF and FF populations. We then calculated $\lambda^{R_1}$, $\lambda^{R_2}$, and the absolute value of the difference between the two (i.e., $\theta$). This was repeated N = 1000 times, after which we determined the proportion of simulations in which $\theta$ was greater than the observed difference between $\lambda^{CF}$ and $\lambda^{FF}$ (i.e., $P[\theta \geq \theta_{obs}]$).

<!-- -   for within stoch and lag: glm/t-test. -->

<!-- see https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12577 for suggestions on text for describing and reporting results -->

```{r glm-lambda}

path <- here("docs/figures/stoch_lambdas.csv")
stoch_lambdas<-read_csv(path)
lambda_glm <- glm(lambda ~ habitat*ipm_type,
             family = gaussian(link = "identity"), data = stoch_lambdas)

# summary(lambda_glm)
# (over)dispersaion
# pearson's residuals
# sum(resid(lambda_glm, type="pearson")^2)/lambda_glm$df.resid
#deviance
# lambda_glm$deviance/model$df.residual
# plot(lambda_glm, which = 1)
# plot(lambda_glm, which = 2)
# https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12577
# acf(residuals(lambda_glm)) ## check autocorrelation
```

Finally, we compared the structure of populations that were projected through the first 250 time steps in each habitat by the IPMs with and without lagged effects. At each time step, we assigned the individuals to one of four stage classes based on plant size, probability of survival, and probability of reproduction (@tbl-plantcats). We then calculated the proportion of the population that was in each stage class in each habitat for each IPM type. Comparing the mean proportion will allow us to infer if two IPMs project different population structures over time, while comparing variances in the proportion will allow us to determine if the population structure for a given IPM is consistent across habitat type. Because Shapiro-Wilk tests indicated that the distributions of proportions were not normally distributed, we used the non-parametric Ansari-Bradley Test to compare the distribution of proportions in each habitat x stage combination. These analyses were conducted using the R packages `rstatix` and `vartest`, respectively [@kassambaraRstatixPipefriendlyFramework2023; @cosarVartestTestsVariance2024].

# Results

```{r }
meta_df <- 
  tar_meta(store = here("_targets")) %>% 
  select(name, size, bytes, time, seconds) %>% 
  filter(str_detect(name, "^ipm_\\w+_[cf]{2}$")) %>% 
  mutate(minutes = seconds / 60,
         hours = minutes / 60)

meta_df_tbl<-meta_df %>% 
  separate(name, into = c("trash", "IPM", "Habitat")) %>% 
  select(-trash) %>% 
  select(IPM, Habitat, minutes) %>% 
  group_by(IPM) %>% 
mutate(IPM = str_replace_all(
    IPM,
    c(
      "det" = "det",  # Deterministic",
      "stoch" = "sk", # Stochastic, kernel-resampled
      "dlnm" = "sp" # Stochastic, parameter-resampled
    ))) %>%
  summarize(mean_time_min = round(mean(minutes), 2)) %>%
  rename("IPM Type" = IPM) %>% 
  rename("mean time (min.)" = mean_time_min)

det_time<-meta_df_tbl %>% filter(`IPM Type`=="det")
det_time<-as.numeric(det_time[1,2])
sker_time<-meta_df_tbl %>% filter(`IPM Type`=="sk")
sker_time<-as.numeric(sker_time[1,2])
spar_time<-meta_df_tbl %>% filter(`IPM Type`=="sp")
spar_time<-as.numeric(spar_time[1,2])

path_det_stats <- here("docs","figures","det_randomization_test.csv")
det_stats<-read_csv(path_det_stats) %>% 
  mutate(p_val=100-perc_below)
```

All IPMs projected higher population growth rates for *Heliconia acuminata* populations in Continuous Forest than for those in Forest Fragments (@tbl-lambdas). The differences between $\lambda^{CF}$ and $\lambda^{FF}$, which ranged from 1.19-2%, were significant for both deterministic ($P[\theta \geq \theta_{obs}]$ = `r det_stats$p_val`, N = 1000 randomization) and stochastic IPMs. There were also significant effects of IPM type, habitat, and their interaction on $\lambda$ (@tbl-glm).

```{r}
stats_table_lambdavars<-stats_table_lambdavars %>% mutate(`p value`=gsub("[*]","",`p value`))
CFstat<-stats_table_lambdavars %>% filter(Comparison=="CF: stochastic vs. lagged") %>% select(Statistic)
FFstat<-stats_table_lambdavars %>% filter(Comparison=="FF: stochastic vs. lagged") %>% select(Statistic)
CFp<-stats_table_lambdavars %>% filter(Comparison=="CF: stochastic vs. lagged") %>% select(`p value`)
FFp<-stats_table_lambdavars %>% filter(Comparison=="FF: stochastic vs. lagged") %>% select(`p value`)
```

Deterministic $\lambda$ was nearly identical to the average $\lambda$ from standard Stochastic IPMs (i.e., without lagged effects, @fig-lambdas). However, including lagged effects in IPMs resulted in lower projections of population growth rate. On average the projections of $\lambda$ from these models were 1.5-2% lower that for the averege projections from stochastic IPMs without lags, while 70-76% of the values were below deterministic $\lambda$ (@fig-lambdas). Including lagged effects in IPMs also significantly increased the variance of projected values of $\lambda$ for both Continuous Forest (C = `r CFstat`, P `r CFp`) and Forest Fragments (C = `r FFstat`, P `r FFp`; @fig-lambdas, @tbl-lambdas). These IPMs nevertheless appear to be more accurate --- the underlying vital rate models for IPMs with lagged effects all had the best fit to the survey data (dAIC = 0, @tbl-aic). <!--- [MU: Describe in methods how we determined model fit] --->

<!--# No longer true if random effect = different smooth for each year -->

The deterministic IPM for Continuous Forest projected slightly more of the smallest and largest plants than the one for Forest Fragments (i.e., the values for these stage classes are below the 1:1 line in [@fig-pop-states]B). This shift towards intermediate-size plants in forest fragments (i.e., pre-reproductive 2) is even more dramatic when using the stochastic IPMs, especially those with lagged effects ([@fig-pop-states]B). However, including lagged effects of climate in IPMs also resulted in more variable projections of population structure ([@fig-pop-states]A, @tbl-meanvars). In continuous forest, the variance in the proportion of the population seedlings and smaller pre-reproductive plants comprise was significantly higher when using IPMs with lagged effects (@tbl-meanvars, @tbl-vartest, @fig-marginal). For forest fragments, however, projecting with lagged effects resulted in significantly higher variance for all five life-history stages (@tbl-vartest).

# Discussion

Mathematical and statistical advances [e.g., @williams2012avoiding; @reesEvolvingIntegralProjection2016; @brooksStatisticalModelingPatterns2019] have rapidly expanded the scope of ecological issues that can be investigated with Integral Projection Models. Our results have three important implications for how these models are used to study one that is increasingly important: the role of climate as a driver of plant demography and population dynamics. The first is that failing to include lagged effects in models could significantly underestimate the growth rate of populations. While relative differences in projected values of $\lambda$ may be sufficient for some tests of theory or experimental comparisons, there are cases in which the specific value projected by models can dramatically shift a study's conclusions. The most obvious of these are when projections of $\lambda$ have implications for conservation or management; even small differences in projections of $\lambda$ can lead to very different conclusions regarding whether or not populations are declining (e.g., $\lambda=0.99$ vs. $\lambda=1.01$) or which management strategy will maximize population growth (e.g., $\lambda=1.01$ vs. $\lambda=1.02$). Given the potential for subtle but significant differences in $\lambda$ between habitats [@brunaDemographicEffectsHabitat2005; @mensahDelayedEffectsWaterlimiting2025], and for lagged effects to either buffer or amplify the effects of climatic variability [@eversInclusionImmediateLagged2023], including these effects in models could be similarly critical for forecasting how plant populations will respond to climate change [@eversInclusionImmediateLagged2023]. This may be especially true for slow-growing and long-lived species like *Heliconia acuminata* [@brunaArePlantPopulations2003; @brunaExperimentalAssessmentHeliconia2002], whose population growth rate may be especially sensitive to temporally varied responses in climate [@eversInclusionImmediateLagged2023].

The second implication is that lagged effects can have important but previously underappreciated consequences for demographic projections beyond the reductions in population growth rate resulting from environmental stochasticity [@doakCorrectlyEstimatingHow2005; @tuljapurkarManyGrowthRates2003; @metcalfStatisticalModellingAnnual2015]. For example, in addition to lower mean $\lambda$ values (@fig-lambdas), models with lagged effects also projected more variable distributions of both $\lambda$ (@fig-lambdas) and population structure (@fig-pop-states). Moreover, these effects were context-dependent: IPMs with lagged effects resulted in more variable projections for both habitat types, but the magnitude of this variation was greater in the forest fragments where lagged effects of precipitation extremes are more pronounced [@scottDelayedEffectsClimate2021].

It has been hypothesized that climate change and habitat fragmentation will synergistic impacts on plant demography [@opdamClimateChangeMeets2004; @lauranceApparentEnvironmentalSynergism2014], and the immediate effects of drought and other climatic extremes on plant mortality are more severe in fragments than continuous forest [@lauranceEffectsStrongDrought2001]. To date, however, only one other study has explicitly considered the potential for synergistic impacts on plant demography of habitat area, the immediate effects of climate, and lagged climatic effects. @hartmannLongtermDataIndicates2023 found that the population growth rate of *Dactylorhiza majalis*, an endangered orchid characteristic of Central Europe's wet grasslands, was strongly related to habitat area, but that the effects of climate were small and ambiguous. We believe our contrasting results are likely explained primarily by differences in the structure of each species' habitat and how it changes with patch area. *Heliconia acuminata* is found in the understory of lowland tropical forest, whose abiotic environment becomes dramatically altered and more variable by the changes in canopy and edge structure following fragment isolation [@didhamEdgeStructureDetermines1999; @brunaEdgeEffectsGrowth2011b]. We hypothesize that lagged effects of climate on demography are indeed likely to be widespread [@eversLaggedDormantSeason2021], they are likely to be strongest in structurally complex environments like tropical forests, where many plant species are already close to their upper thermal limits [@sentinellaTropicalPlantsNot2020], rather than more open and hence climatically variable ones such as temperate grasslands. Assessing this hypothesis requires researchers to not only heed recent calls to focus effort on how climate influences demography [@eversInclusionImmediateLagged2023; @williamsLinkingClimateDemography2025], but to do so with populations found across a range of habitats and ecological contexts.

<!-- In other words, populations in forest fragments had proportionately less recruitment and fewer individuals growing into the larger, reproductive size classes [@brunaDemographicEffectsHabitat2005].  -->

<!-- We hypothesize that the more skewed structure of *H. acuminata* populations projected for fragments, which mirrors that documented across the BDFFP reserves [@brunaHabitatFragmentationDemographic2002], is probably due to the delayed effects of supra-annual climatic extremes compounding the chronic reductions in seedling recruitment [@brunaEffectsForestFragmentation2002; @brunaSeedGerminationRainforest1999] and plant growth [@gagnonGrowthUnderstoryHerb2011] observed in fragments. -->

The final implication is that computational costs may hinder the widespread consideration of lagged effects for the foreseeable future --- even for studies of a single species. Our analyses were conducted on the University of Florida's HiperGator supercomputer using one nVidia A100 GPU with an allocation of 8 GB RAM. While each iteration of the Deterministic IPM and Stochastic IPM without lagged effects took only \~`r det_time` and \~`r sker_time` min to iterate, respectively, an iteration of the Stochastic IPM with lagged effects took \~`r spar_time` min. This is largely due to the required computational resources and algorithms (i.e., `predict()`), which are much slower for Generalized Additive Models with 2D smooths than for the GAMs used in Deterministic and Stochastic IPMs. Advances in computational power and access to supercomputer could lower the temporal and computational cost of running the thousands of iterations required to estimate $\lambda$ and its confidence intervals, as could GAMs for large datasets [i.e., @mgcv]. Until these resources are more broadly available, however, it will be important to consider if any model assumptions can be relaxed to reduce the "computational burden" [*sensu* @eversInclusionImmediateLagged2023] or when including lagged effects in demographic models might be truly essential or beneficial. Indeed, @compagnoniAntecedentEffectModels2024 found that using highly sophisticated antecedent effect models [@ogleQuantifyingEcologicalMemory2015a] did not significantly improve the ability to predict the effects of climate on vital rates, though this may have been due to climate signals that were too weak or noisy to influence the demography of their primarily temperate focal species. <!--- [MU: Are there optimization possibilities with mgcv::bam?] EB: added last sentence to paragraph above but note eric's comment: can't really be paralellized because each iteration depends on the previous one.  There's *maybe* some potential speed gains in using multiple cores with predict.bam(), but unlikely to be substantial. -->

It is important to emphasize, however, that computational power cannot compensate for limited data. Detecting lagged effects of climate and evaluating their consequences requires long-term demographic data --- data that are only available for a relatively small number of species, of which few are in the tropics [@brunaHabitatFragmentationPlant2009; @romerPlantDemographicKnowledge2024] and where the synergistic effects of climate- and land-use change may be particularly acute [@lauranceApparentEnvironmentalSynergism2014; @opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015]. The increasing evidence that lagged effects are ubiquitous, and that they can have major demographic impacts, underscores the need to support the collection of such long-term data, the complementary development of experimental and statistical approaches to disentangling lagged effects, and community driven efforts to identify priority or model systems for in-depth investigation.\

<!-- ### Acknowledgments {.unnumbered} -->

<!-- We thank M. Acevedo, O. Acevedo-Charry, and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript. We also thank Sam Levin for his help with the `ipmr` package. Financial support for collecting the demographic data used in this study was provided by the U.S. National Science Foundation (awards INT 98-06351, DEB-0309819, DEB-0614149,DEB-0614339, and DEB-1948607). This article is publication no. \_\_\_ in the BDFFP Technical series. The authors declare no conflicts of interest. -->

<!-- ### CRediT Statement {.unnumbered} -->

<!-- <!-- https://casrai.org/credit/ -->

<!-- The authors made the following contributions: Eric R. Scott: conceptualization, methodology, formal analysis, validation, visualization, data curation, software, writing – original draft, review, and editing; Emilio M. Bruna: conceptualization, methodology, formal analysis, validation, visualization, data curation, software, writing (original draft, review and editing), funding acquisition, project administration, supervision; Maria Uriarte: conceptualization, methodology, writing (review and editing), funding acquisition. -->

### Data Availability Statement {.unnumbered}

Data and R code used in this study are archived with Zenodo at *(doi and url to be added on acceptance)*. Anonymized versions are available to Editors and Reviewers at <https://osf.io/mw5a7/overview?view_only=f0ea421ee916437eac32d59debfdc894>

<!-- ```{r } -->

<!-- pkgs <- cite_packages(output = "table", out.dir = ".") -->

<!-- knitr::kable(pkgs) -->

<!-- ``` -->

\newpage

# References {.unnumbered}

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

<!-- # Figures -->

\newpage

```{r #fig-lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Life cycle diagram of *Heliconia acuminata*. Each transition is associated with an equation for a vital rate function.  The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM. The table below shows the equivalent equations for stochastic IPMs with and without lagged effects (i.e., kernel-resampled vs. parameter-resampled IPMs)."

# Draft of this figure is in docs/figures/Heliconia lifecycle DAG.key
path <- here("docs","figures","lifecycle.png")
knitr::include_graphics(path, rel_path = TRUE)

```

```{r #fig-lambdas,  echo=FALSE, out.width="100%",fig.align='center'}
#| fig.cap = "Distribution of 900 values of $\\lambda$ projected with (A) Stochastic IPMs without lagged effects and (B) Stochastic IPMs with lagged effects. IPMs were used to project $\\lambda$ for both Continuous Forest (above, in green) and Forest Fragments (below, in gray). The solid line indicates the mean value of $\\lambda$, the dashed line indicates the value of $\\lambda$ in that habitat projected with Deterministic IPMs."

# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 

path <- here("docs","figures","lambdas_fig.png")
knitr::include_graphics(path, rel_path = TRUE)
```

```{r #fig-pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "(A) The change over time in the the proportion of _Heliconia acuminata_ populations in different size/stage classes when simulating population dynamics in Continuous Forest (CF) or Forest Fragment (FF) with three different integral projection models (e.g., IPMs). Results are shown for the first 250 iterations of populations; for the criteria used to define the size categories see @tbl-plantcats. (B) The relative proportion of the population in each size class (FF vs. CF) for 250 iterations of each IPM model. Note that this excludes transient dynamics (iterations 1-30). Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class."
# note flowering prob for rp1 <0.25 but for rp2>0.2???'
# include _seedlings_ (a discrete category in the IPMs), _pre-reproductive 1_ (log(size) = 0–2.5, average survival < 0.9, probability of flowering \\approx 0), _pre-reproductive 2_ (log(size) = 2.5–4.5, average survival probabilty > 0.8, probability of flowering \\approx 0), _reproductive 1_ (log(size) = 4.5–6, average survival probability >0.95, flowering probability < 0.25), and _reproductive 2_ (log(size) = 6+, average survival probability >0.95, flowering probability > 0.2)
knitr::include_graphics(here(fig_pop_states), rel_path = TRUE)
```

```{r #fig-marginal,  echo=FALSE, out.width="75%",fig.align='center'}
#| fig.cap = "The relative proportion of the population in each size class (FF vs. CF) for each of 250 iterations of the no-lag IPMs (dark shading) and Stochastic IPMs with lagged effects (light shading). Values on the 1-1 line indicate an iteration where CF and FF have the same proportion of the population in a given size class; the marginal plots indicate the distribution of these relative proportions for each class of IPM in each habitat. Note that both the scatterplots and marginal plots exclude transient dynamics (iterations 1-30)."

# The figures is currently generated with "plotting_lambdas.R, need to move to targets_r 

path <- here("docs","figures","marginal_stage_fig.png")
knitr::include_graphics(path, rel_path = TRUE)
```

\newpage

<!-- # Tables -->

```{r}
library(pander)
library(glue)
library(snakecase)
```

```{r tbl-plantcats, results='asis'}
#| label: tbl-plantcats
#| tbl-cap: "Size and stage categories used for comparing _Heliconia acuminata_ population structure. Note that seedlings are a discrete size class not based on size (see _Methods_ for additional details)."
#| tbl-pos: H

category<-c("Seedlings", "Pre-reproductive 1", "Pre-reproductive 2", "Reproductive 1", "Reproductive 2")
size<-c(" - ","0–2.5","2.5-4.5","4.5-6","$\\geq$ 6")
survival<-c("-","$\\leq$ 0.9","$\\geq$ 0.8","$\\geq$ 0.95","$\\geq$ 0.95")
flowering<-c("-","$\\approx$ 0","$\\approx$ 0","$\\leq$ 0.25","$\\geq$ 0.2")
cat_table<-bind_cols(Category=category,
                     `Log(size)`=size, 
                     `Avg. prob. survival` = survival,
                     `Prob. flowering` = flowering)
cat_table %>% 
kable(digits = 2,
  format = "latex",
  align = "lccc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>% 
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) 
# %>% 
#   column_spec(1, width = "50em") %>% 


# cat_table %>% 
#   pandoc.table(
#   digits = 5,
#   keep.trailing.zeros = TRUE,
#   justify = c("left", "center", "center", "center"),
#   full_width = FALSE
# )

```

\newpage

```{r tbl-lambdas, results='asis'}
#| label: tbl-lambdas
#| tbl-cap: "Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
#| tbl-pos: H
options(scipen = 999)
lambda_vars<-lambda_vars %>% 
  rename(ipm=ipm_type) %>% 
  mutate(ipm=if_else(ipm=="lag","dlnm",ipm))

lambda_table %>% 
  left_join(lambda_vars,by=c("ipm","habitat")) %>% 
  mutate(habitat = toupper(habitat)) %>% 
  mutate(across(c(est, lower, upper), ~format(round(.,4)))) %>% 
  # mutate(lambda = glue::glue("{est} ({lower}, {upper})")) %>%
  mutate(CI = glue::glue("({lower}, {upper})")) %>%
  mutate(ipm = str_replace_all(
    ipm,
    c(
      "det" = "Deterministic",
      "stoch" = "Stochastic no lag",
      "dlnm" = "Stochastic lagged"
    )
  )) %>% 
  # separate(lambda,c("lambda","CI"),sep=" [(]") %>% 
  select(IPM = ipm, 
         Habitat = habitat, 
         # "$\\lambda$" = lambda) %>% 
         "$\\lambda$" = est,
         "95\\% CI (Lower, Upper)"=CI,
         Variance=variance) %>%
  mutate(Variance=as.character(Variance)) %>% 
  mutate(Variance=if_else(IPM=="Deterministic", "-", Variance)) %>% 
  mutate(IPM = str_replace_all(
    IPM,
    c(
      "Deterministic" = "",
      "Stochastic no lag" = "",
      "Stochastic lagged" = ""
      )
  )) %>% 
  kable(digits = 2,
  format = "latex",
  align = "lcccc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>% 
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>% 
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "15em") %>% 
  pack_rows("Deterministic", 1, 2) %>%
  pack_rows("Stochastic", 3, 4) %>% 
  pack_rows("Stochastic with lagged effects", 5, 5) 
  
# 
# pandoc.table(
#   digits = 5,
#   keep.trailing.zeros = TRUE,
#   justify = c("left", "center", "center"),
#   full_width = FALSE
  # ,
  # caption = "(\\#tab-lambdas) Population growth rates for continuous forest (CF) and forest fragments (FF) under different kinds of IPMs with bootstrapped, bias-corrected, 95% confidence intervals."
# )

```

\newpage

```{r table-glm, results="asis"}
#| label: tbl-glm
#| tbl-cap: "Estimated parameters, standard errors, t-values and P-values for the GLM of the effect of Habitat and IPM Type on projections of lambda."
#| tbl-pos: H

glm_table<-tidy(lambda_glm) %>% 
  mutate(estimate=round(estimate,3),
  std.error=round(std.error,4),
  statistic=round(statistic,3),
  p.value=round(p.value,4)) %>% 
  mutate(p.value=if_else(p.value<0.001,as.character("< 0.0001"),as.character(p.value))) %>% 
  # mutate(p.value=if_else(p.value< 0.01,"0.001",p.value)) %>% 
  rename(Term=term,
         Estimate=estimate,
         SE=std.error,
         "z value"=statistic,
         "\\textit{P}"=p.value) %>% 
  mutate(Term=gsub("[:]"," : ",Term)) %>%
  mutate(Term=gsub("habitatff","$Habitat_{FF}$",Term)) %>%
  mutate(Term=gsub("ipm_typestoch","$IPM_{Stoch}$",Term)) 

# %>%
#   mutate(Term=gsub(":"," ",Term)) %>% 
#   mutate(Term=gsub("_"," ",Term)) 

glm_table %>% 
  kable(
  digits = 2,
  format = "latex",
  align = "lccccc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  )

```

\newpage

```{r tbl-vartest, results='asis'}
#| label: tbl-vartest
#| tbl-cap: "Results of statistical tests comparing the variance in the proportion of each habitat's population projected to be in each stage class by Stochastic IPMs with and without lagged effects (N = 220 projections per stage class).  The variances for each habitat x stage class x IPM combination can be found in @tbl-meanvars. Comparisons where the p-value of the test was < 0.05 are indicated with an asterisk."
#| tbl-pos: H

# source(here("R","variance_stats.R"))
# variance_stats()
path <- here("docs","figures","varstats_table.csv")
stats_table<-read_csv(path)

stats_table %>%
  # select(-`Vital Rate`) %>%
  rename(`Stage`=`Vital Rate`,
         "\\textit{df}"=df,
         "\\textit{P}"=`p value`) %>%  
  mutate(`Stage`= " ") %>% 
  # mutate(Comparison="KR v PR") %>% 
  # relocate(Comparison,.before=1) %>% 
kable(digits = 5,
  format = "latex",
  align = "lcrrcr",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  # caption = "Test comparing variances of vital rates in stochastic ipms.",
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>%
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "8em") %>%
  # collapse_rows(columns = 1:2,valign="top")
  pack_rows("Seedling", 1, 2) %>%
  pack_rows("Pre-reproductive 1", 3, 4) %>%
  pack_rows("Pre-reproductive 2", 5, 6) %>%
  pack_rows("Reproductive 1", 7, 8) %>%
  pack_rows("Reproductive 2", 9, 10)

```

\newpage
\appendix
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{figure}{0}
\setcounter{table}{0}

# Supplementary Materials {.appendix .unnumbered}

```{r aic, results='asis'}
#| label: tbl-aic
#| tbl-cap: "Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic IPMs without lagged effects (i.e., kernel-resampled models); and those with a distributed lag non-linear model (DLNM) were used to build stochastic IPMs with lagged effects (i.e., parameter-resampled models). <i>'edf'</i> is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
#| tbl-pos: H
#| 
aic_tbl_df %>%
  group_by(habitat, model) %>% 
  # mutate(rank = row_number()) %>% 
  ungroup() %>% 
  mutate(habitat = toupper(habitat),
         env_effect_incl = ifelse(
           env_effect_incl != "DLNM",
           to_sentence_case(as.character(env_effect_incl)),
           as.character(env_effect_incl)
         )
  ) %>% 
  mutate(
    model = fct_relevel(
      model,
      c("surv", "size", "flwr", "surv_sdlg", "size_sdlg")
    ),
    model = fct_recode(
      model,
      "Survival"          = "surv",
      "Growth"            = "size",
      "Flowering"         = "flwr",
      "Seedling survival" = "surv_sdlg",
      "Seedling growth"   = "size_sdlg"
    )
  ) %>% 

   #another option is to use psuedo model notation.  Can't fit both unless I remove the "habitat" column or something.
  mutate(RHS = str_replace_all(
    env_effect_incl,
    c(
      "None" = "~s(z)",
      "Random effect of year" = "~s(z, by = 'year')",
      "DLNM" = "~s(z) + te(spei, lag)"
    )
  )) %>% 
  
  arrange(habitat, model) %>% 
  mutate(habitat=c(rep("Continuous Forest",15),rep("Forest Fragments",15))) %>% 
  mutate(model=(rep(c("Survival","","","Growth","","","Flowering","","","Seedling survival","","","Seedling growth","",""),2)))  %>%
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "\\textit{edf}" = "df",
    "\u0394AIC" = "dAIC"
  )  %>% 
  mutate(Habitat = str_replace_all(
    Habitat,
    c(
      "Continuous Forest" = "",
      "Forest Fragments" = ""
      )
  )) %>% 
  kable(digits = 2,
  format = "latex",
  align = "cllcc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  ) %>% 
# %>% 
  # collapse_rows(columns = 1:2, valign = "top")
  column_spec(1, width = "10em") %>% 
  pack_rows("Continuous Forest", 1, 15) %>%
  pack_rows("Forest Fragments", 16, 30)



# pandoc.table(
#   justify = c( "left", "left","left", "right",  "right"),
#   full_width = FALSE
#   # ,
#   # caption = Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
# )


```

\newpage

::: landscape
<!-- \renewcommand{\arraystretch}{1} -->

```{r tbl-meanvars, results='asis'}
#| label: tbl-meanvars
#| tbl-cap: "Summary statistics (median, mean, and variance) describing the proportion of populations projected to be in each of five life-history stages by Stochastic IPMs with and without lagged effects (i.e., kernel- vs. parameter-resampled IPMs, N = 220 projections for each IPM class x habitat combination.)"
#| tbl-pos: H

# source(here("R","variance_stats.R"))
# variance_stats()
#array stretch increases row height
cat("\\renewcommand{\\arraystretch}{0.8}   \n")
path <- here("docs","figures","meanvar_table.csv")
options(scipen = 999)
meavvar_table<-read_csv(path) %>%
  select(-n) %>% 
  mutate(meanCF=round(meanCF,4)) %>% 
  mutate(meanFF=round(meanCF,4)) %>% 
  mutate(medianCF=round(medianCF,4)) %>% 
  mutate(medianFF=round(medianFF,4)) %>% 
  mutate(varCF=round(varCF,4)) %>% 
  mutate(varFF=round(varFF,4)) %>% 
  mutate(varCF=if_else(varCF<0.0001,as.character("< 0.0001"),as.character(varCF))) %>% 
  mutate(varFF=if_else(varFF<0.0001,as.character("< 0.0001"),as.character(varFF))) %>% 
  rename(Stage=log_size_bin) %>% 
  mutate(Stage=" ") 
  
names(meavvar_table)<-c("Stage","IPM","CF","FF","CF","FF","CF","FF")
# %>% 
  # mutate(CF=paste(meanCF,"±",varCF,sep=" "),.keep="unused") %>% 
  # mutate(FF=paste(meanFF,"±",varFF,sep=" "),.keep="unused")
# pandoc.table(meavvar_table,
#   digits = 5,
#   keep.trailing.zeros = TRUE,
#   justify = c("left", "center", "center","center","center"),
#   full_width = FALSE,
# caption = "(\\#tab-meanvars) caption."
# )
meavvar_table %>%
kable(
  # digits = 2,
  # format = "latex",
  align = "llcccccc",
  escape = FALSE,
  row.names = FALSE,
  booktabs = T,
  # caption="Summary statistics (median, mean, and variance) describing the proportion of projected populations in each of five life-history stages (N = 220 projections for each class of stochastic IPM).",
  linesep = ""
) %>%
  kable_styling(
    bootstrap_options = c("hover"),
    # full_width = F,
    latex_options = c("scale_down"),
    # latex_options = c("hold_position"),
    font_size = 12,
    position = "center"
  )%>%
  # collapse_rows(columns = 1:2, valign = "top")
  # column_spec(1, width = "15em") %>%
  add_header_above(c(" " = 2, "Median" = 2, "Mean" = 2, "Variance" = 2)) %>%
  column_spec(1, width = "8em") %>%
  # column_spec(7, width = "6em") %>%
  # column_spec(8, width = "6em") %>%
  pack_rows("Seedling", 1, 2) %>%
  pack_rows("Pre-reproductive 1", 3, 4) %>%
  pack_rows("Pre-reproductive 2", 5, 6) %>%
  pack_rows("Reproductive 1", 7, 8) %>%
  pack_rows("Reproductive 2", 9, 10)
# cat("\\renewcommand{\\arraystretch}{1}   \n")
```
:::
