---
title: "Full Title"
shorttitle: "Running Title"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF_WEC]
      correspondence: true
  - Co-author two:
      email: secondperson@university.edu
      institute: [affiliation2]
      correspondence: false
  - Co-author three:
      email: thirdperson@university.edu
      institute: [UF_WEC, affiliation2]
      correspondence: false
      
institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - affiliation2: Department of Ecology, Evolution and Environmental Biology, Example University 123 Center Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      #- -Fpandoc-crossref #required for cross-ref equations
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
keywords: |
  keyword1; keyword2; keyword3
highlights: |
  These are the highlights
  
editor_options:
  markdown:
    wrap: sentence
    canonical: true
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Running title: `r rmarkdown::metadata$shorttitle`

<!-- Highlights: `r rmarkdown::metadata$highlights` -->

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
```

<!-- The actual document text starts here: -->

\pagebreak

# Abstract

Abstract text.

\pagebreak

# Introduction

Introduction with an in-text citation [@scottInteractiveEffectsDrought2019].

# Methods

Methods text.

# Results

A reference to Table \@ref(tab:example-table).
A reference to Figure \@ref(fig:example-figure).

# Discussion

# Acknowledgments

We thank **,** \_, \_\_\_ and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
Financial support was provided by the U.S.
National Science Foundation (awards \_\_\_\_, and \_\_\_\_).
The authors delcare no conflicts of interest.

# Data Availability Statement

Data and R code used in this study are archived with Zenodo at <doi url>.

\pagebreak

# Tables

```{r}
library(pander)
```

```{r example-table, results='asis'}
pandoc.table(
  head(iris),
  justify = c("right", "right", "right", "right", "left"),
  full_width = FALSE,
  caption = "(\\#tab:example-table) A table must include the table tag in order to cross-reference it correctly."
)
```

# Figures

```{r}
#| example-figure,
#| fig.cap = "A caption for an example figure. It can be long and span mutiple
#|   lines like this."
plot(cars)
```

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::
